### 2014-08-29 paper/paperLuke.Rmd
test commit to pull/commit
tsting

Seven Things That Affect Education Around the World
========================================================

@@

The OECD describes the PISA survey as ``the world's global metric for quality, equity and efficiency in school education" (citep[p. 3]{PISA2014a}). The goal of the PISA survey is to assess the workforce readiness of 15-year old students. Nearly 500,000 students were tested across the 65 countries and economies. Students are examined on how well they can apply the knowledge they learned in school to applications outside of school. The scores reported range between 0-1000. Additional information about the students, parents, and schools is also collected. The students completed a questionnaire providing information about themselves, their homes, their schools, and a variety of psychological views regarding factors they believe affects their performance in school. Schools principals responded to a questionnaire covering the school system and learning experiences for students. In some countries, parents completed a questionnaire requesting information about their perceptions regarding the school system, expectations for their child, and their involvement in their child's schooling.

The data was made available by the OECD as part of a data challenge for the useR! 2014 conference. Entries to the competition can be found at http://beta.icm.edu.pl/PISAcontest/. We learned a lot from this data, and won one of the prizes. This is a description of the seven main things we learned about 15 year olds, their math, science and reading skills, and the effects of other factors.

@@

```{r load_packages, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
library(dplyr)
library(stringr)
library(reshape2)
library(ggplot2)
library(lubridate)
library(ggvis)
library(maps)
library(ggmap)
library(htmltools)
library(rworldmap)
library (grid)    
library (scales)
library(doBy)
```

```{r load_data, echo = FALSE, message = FALSE, warning = FALSE, results='hide', cache=TRUE}
# This is the scaffolding to import the data
sets <- c("item", "parent", "school", "scoredItem", "student")

@@

world <- getMap(resolution = "low")
library(plyr)
world.polys <- extractPolygons(world)
detach("package:plyr")
```

*** We should decide on country names, to coordinate map info and the test data. This si some really ugly code to convert names to be the same between the two. Could be made cleaner, much cleaner!

```{r dataprep, echo = FALSE, message = FALSE, warning = FALSE}
colnames(student2012)[1] <- "name"
student2012$name <- as.character(student2012$name)
colnames(school2012)[1] <- "name"
school2012$name <- as.character(school2012$name)

fix_country = function(df) {
df$name = as.character(df$name)
df$name[df$name=="Serbia"] <- "Republic of Serbia"
df$name[df$name=="Korea"] <- "South Korea"
df$name[df$name=="Chinese Taipei"] <- "Taiwan"
df$name[df$name=="Slovak Republic"] <- "Slovakia"
df$name[df$name=="Russian Federation"] <- "Russia"
df$name[df$name=="Perm(Russian Federation)"] <- "Russia"
df$name[df$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
df$name[df$name=="China-Shanghai"] <- "China"
df$name[df$name=="China-Macau"] <- "China"
df$name[df$name=="Connecticut (USA)"] <- "United States of America"
df$name[df$name=="Florida (USA)"] <- "United States of America"
df$name[df$name=="Massachusetts (USA)"] <- "United States of America"

df
}

# Now add data
#student2012$name[student2012$name=="Serbia"] <- "Republic of Serbia"
#student2012$name[student2012$name=="Korea"] <- "South Korea"
#student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
#student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
#student2012$name[student2012$name=="Russian Federation"] <- "Russia"
#student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
#student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong S.A.R."
#student2012$name[student2012$name=="China-Shanghai"] <- "China"
#student2012$name[student2012$name=="China-Macau"] <- "China"
#student2012$name[student2012$name=="Connecticut (USA)"] <- "United States of America"
#student2012$name[student2012$name=="Florida (USA)"] <- "United States of America"
#student2012$name[student2012$name=="Massachusetts (USA)"] <- "United States of America"
student2012$ST06Q01 <- as.numeric(student2012$ST06Q01)
student2012$PV1MATH <- as.numeric(student2012$PV1MATH)
student2012$PV1READ <- as.numeric(student2012$PV1READ)
student2012$PV1SCIE <- as.numeric(student2012$PV1SCIE)
student2012$SENWGT_STU <- as.numeric(student2012$SENWGT_STU)
school2012 = fix_country(school2012)
student2012 = fix_country(student2012)
```

### Teachers, computers, and libraries at school make a difference in math scores

```{r shortages, echo = FALSE, message = FALSE, warning = FALSE}
school2012.sub = school2012[, c(1, 6, 34, 35, 40, 45, 46, 47, 48, 49, 50, 51)] #%>% select(name, SCHOOLID, SC11Q02, SC11Q03, SC14Q02, SC14Q07, SC14Q08, SC14Q09, SC14Q10, SC14Q11, SC14Q12, SC14Q13)
student2012.sub = student2012[, c(1, 6, 501, 541, 546, 634)] #%>% select(name, SCHOOLID, PV1MATH, PV1READ, PV1SCIE)

scores_by_school = summarise(group_by(student2012.sub, name, SCHOOLID), mmath = mean(PV1MATH), 
                                      mread = mean(PV1READ), msci = mean(PV1SCIE))

df = scores_by_school %>% left_join(school2012.sub)
df2 <- summarise(group_by(df, name, SC14Q02), mmath2 = mean(mmath))
qplot(SC14Q02, mmath2, data = df2, facets = ~name)
df3 <- summarise(group_by(df, name, SC14Q10), mmath2 = mean(mmath))
qplot(SC14Q10, mmath2, data = df3, facets = ~name)
```

### The gender gap in math is not universal but the reading gap is, in favor of girls

For each country, average scores on math and reading are calculated separately for boys and girls. These averages are differenced to compute the "gap" between boys and girls. T-statistics were calculated to determine the statistical significance of the difference. The results are plotted in the graphs below. Each dot corresponds to a difference between girls and boys scores, for each country. Color indicates statistical significance (blue in favor of boys, pink in favor of girls and green means there was no significant difference). The size of the dot indicates the proportion of boys in the sample size. For most countries it is roughly equal, but some countries have an imblance, for example, Malaysia has more girls tested than boys. The maps color the countries by the presence of a gender gap.

The countries are sorted from biggest gap to smallest gap. Most countries have a gender gap in math scores in favor of boys. Colombia has the biggest gender gap, about 30 points on average boys do better than girls. It should be noted that this is small, though, because this 30 points is out of a possible 1000 points.  Four countries have gender gaps in favor of girls, and even more surprising, these are middle eastern, or Muslim countries.

Conversely, though, when we look at reading scores the story is entirely different. Universally, girls score significantly better than boys. Again, the size of the gap is small, between 10 and 80 points.

Although, not shown here, it also should be noted that this is the difference between the average scores for boys and girls. There is a slightly different story for each country if we look at the top math score for each geneder only, where we would notice that the top score in the USA is achieved by a girl, and is substantially higher than the highest boys score. There is far more individual variability than the overall average differences.

```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE}
student2012.sub <- student2012[, c(1, 12, 501, 541, 546, 634)]

@@

student2012.sub.summary.gap <- summarise(group_by(student2012.sub, name), 
                  mathgap=mean(PV1MATH[ST04Q01=="Male"], na.rm=T)-
                          mean(PV1MATH[ST04Q01=="Female"], na.rm=T),
                  wmathgap=weighted.mean(PV1MATH[ST04Q01=="Male"], w=SENWGT_STU[ST04Q01=="Male"], na.rm=T)-
                          weighted.mean(PV1MATH[ST04Q01=="Female"], w=SENWGT_STU[ST04Q01=="Female"], na.rm=T),
                  mtest.stat = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$statistic, 
                  mp.value = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$p.value,
                  readgap=mean(PV1READ[ST04Q01=="Male"], na.rm=T)-mean(PV1READ[ST04Q01=="Female"], na.rm=T),
                  rtest.stat = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$statistic, 
                  rp.value = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$p.value,
                  sciencegap=mean(PV1SCIE[ST04Q01=="Male"], na.rm=T)-mean(PV1SCIE[ST04Q01=="Female"], na.rm=T),
                  stest.stat = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$statistic, 
                  sp.value = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$p.value,                                         
                  minmale=min(PV1MATH[ST04Q01=="Male"], na.rm=T), 
                  minfemale=min(PV1MATH[ST04Q01=="Female"], na.rm=T), 
                  maxmale=max(PV1MATH[ST04Q01=="Male"], na.rm=T), 
                  maxfemale=max(PV1MATH[ST04Q01=="Female"], na.rm=T), 
                  propmale=length(PV1MATH[ST04Q01=="Male"])/length(PV1MATH), 
                  propfemale=length(PV1MATH[ST04Q01=="Female"])/length(PV1MATH))
#qplot(mathgap, wmathgap, data=student2012.sub.summary.gap, xlab="Mean", ylab="Weighted Mean",
#      xlim=c(-30,30), ylim=c(-30,30)) + geom_abline(slope=1) + theme(aspect.ratio=1)
student2012.sub.summary.gap$msig <- ifelse(student2012.sub.summary.gap$mp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat>0] <- "male"
student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat<0] <- "female"
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$mathgap)])
qplot(name, mathgap, data=student2012.sub.summary.gap, size=propmale, color=msig) + 
  xlab("") +
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5)) + 
  scale_size("Prop male") +
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$maxmale)])
#qplot(name, maxmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
#  xlab("") +
#  geom_point(aes(y=maxfemale), colour="pink") + 
#  scale_y_continuous("High Math") + coord_flip() + theme_bw()
#ggsave("gendermathtop.pdf", width=3.5, height=7)
#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$minfemale)])
#qplot(name, minmale, data=student2012.sub.summary.gap, color=I("skyblue")) + 
#  xlab("") +
#  geom_point(aes(y=minfemale), colour="pink") + 
#  scale_y_continuous("Low Math") + coord_flip() + theme_bw()
#ggsave("gendermathbottom.pdf", width=3.5, height=7)
#
student2012.sub.summary.gap$rsig <- ifelse(student2012.sub.summary.gap$rp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat>0] <- "male"
student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat<0] <- "female"
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$readgap)])
qplot(name, readgap, data=student2012.sub.summary.gap, size=propmale, color=rsig) + 
  xlab("") +
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Reading Score Gap") + 
  scale_size("Prop male") +
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
#ggsave("genderreadgap.pdf", width=5, height=8)
#student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"
#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
#qplot(name, sciencegap, data=student2012.sub.summary.gap, size=propmale, color=ssig) + 
#  xlab("") +
#  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
#  scale_y_continuous("Science Score Gap") + 
#  scale_size("Prop male") +
#   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()

@@

```{r gendermaps, fig.width=12, fig.height=7, warning=FALSE, message=FALSE, echo=FALSE}
student2012.sub.map <- left_join(student2012.sub.summary.gap, world.polys)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=msig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
#ggsave("gendermathmap.pdf", width=12, height=7)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=rsig)) +
  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  new_theme_empty + theme(legend.position="none")
#ggsave("genderreadmap.pdf", width=12, height=7)
#ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
#  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=ssig)) +
#  scale_fill_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
#  new_theme_empty + theme(legend.position="none")
```

### Starting school at an older age generated lower average math score.

For each country we have computed the average math test score by age that the children started school. In additon, for each country we computed the median age that children in the study started school. There are differences in typical school start age between countries. Ireland has the youngest median start age - children head to school at age 4. Britain, and its past colonies, Australia, Canada and New Zealand, have a median start age of 5. For many other countries the median start age is 6, and for northern Europe, much of Asia and Brazil median start age is as late as 7. Some countries appear to have rigid starting ages, eg in Japan all children start at age 6.

```{r age, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]
student2012.sub.age <- summarise(group_by(student2012.sub, name), 
                  math4 = mean(PV1MATH[ST06Q01 == 4], na.rm = T), 
                  math5 = mean(PV1MATH[ST06Q01 == 5], na.rm = T), 
                  math6 = mean(PV1MATH[ST06Q01 == 6], na.rm = T), 
                  math7 = mean(PV1MATH[ST06Q01 == 7], na.rm = T), 
                  math8 = mean(PV1MATH[ST06Q01 >  7], na.rm = T))
student2012.sub.age.m <- melt(student2012.sub.age)
student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
student2012.sub.agemedian <- summarise(group_by(student2012.sub, name), medianstart = median(ST06Q01, na.rm=T))
orderedbymedianage <- student2012.sub.agemedian$name[order(student2012.sub.agemedian$medianstart)]
student2012.sub.age.m.median <- left_join(student2012.sub.age.m, student2012.sub.agemedian)
student2012.sub.age.m.median$medianstart <- factor(student2012.sub.age.m.median$medianstart)
student2012.sub.age.m.median$name <- factor(student2012.sub.age.m.median$name, levels=orderedbymedianage)
qplot(variable, value, data=student2012.sub.age.m.median, xlab="Age at start", ylab="Math Scores", 
      geom=c("point","smooth"), se=F, group=name, colour=medianstart) + 
 facet_wrap(~name, ncol=8) + theme_bw() + theme(legend.position="none")
# ggsave("school-start.pdf", width=14, height=14)
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  read4 = mean(PV1READ[ST06Q01 == 4], na.rm = T), 
#                  read5 = mean(PV1READ[ST06Q01 == 5], na.rm = T), 
#                  read6 = mean(PV1READ[ST06Q01 == 6], na.rm = T), 
#                  read7 = mean(PV1READ[ST06Q01 == 7], na.rm = T), 
#                  read8 = mean(PV1READ[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Reading Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  scie4 = mean(PV1SCIE[ST06Q01 == 4], na.rm = T), 
#                  scie5 = mean(PV1SCIE[ST06Q01 == 5], na.rm = T), 
#                  scie6 = mean(PV1SCIE[ST06Q01 == 6], na.rm = T), 
#                  scie7 = mean(PV1SCIE[ST06Q01 == 7], na.rm = T), 
#                  scie8 = mean(PV1SCIE[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Science Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 
```

```{r agemaps, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
# Extract map polygons for modern world
world <- getMap(resolution = "low")
library(plyr)
world.polys <- extractPolygons(world)
detach("package:plyr")

student2012.sub.agemedian$medianstart <- factor(student2012.sub.agemedian$medianstart)
student2012.sub.map <- left_join(student2012.sub.agemedian, world.polys)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=medianstart)) +
  scale_fill_discrete("Median Start Age") +
  new_theme_empty + theme(legend.position="bottom")
#ggsave("agestartmap.pdf", width=12, height=7)
```
### Time spent out of school studying is important, but only up to a point

***Need to get this code working

```{r, eval=FALSE}
# Out of School Study time
b1 <- summaryBy(data=student2012, pvM + pvR + pvS+ ST57Q01+ ST57Q02 +ST57Q03+ ST57Q04+ ST57Q05+ ST57Q06 ~CNT,FUN=mean, na.rm=T)
names(b1) <- c("CNT","pvM","pvR","pvS","ST57Q01", "ST57Q02","ST57Q03", "ST57Q04", "ST57Q05", "ST57Q06" )
# Out of school study time (total hours HW)
b1$CNT <- reorder(factor(b1$CNT),b1$pvS, mean)
b2 <- melt(b1[,c("CNT","pvM","pvR","pvS")],id="CNT")
b2 <- cbind(b2, b1[match(b2$CNT,b1$CNT), 5:10])
qplot(CNT, value,color=variable,size=ST57Q01, data = b2)+ coord_flip()
head(b2)
qplot(ST57Q01, value,color=variable,size=ST57Q02, data = b2)+ coord_flip()+facet_wrap(~CNT)

student.sub <- studschool[,c("name","pvM","pvR","pvS","ST57Q01","OECD","SC03Q01")]
study <- melt(student.sub,id=c("name","ST57Q01","OECD","SC03Q01"))
names(study) <- c("Country","HomeworkHours","OECD","AREA","Subject","Score")
levels(study$Subject) <- c("Math","Reading","Science")
p <- ggplot(data = study, aes(HomeworkHours,Score,colour=Subject) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)+ xlab("Out-of-School Study Hours")
#ggsave(p,file="studytime.pdf",height=13,width=13)
p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") +xlim(0,30)
ggsave(p,file="HistHWschool.pdf",width=8,height=6)

student.sub <- studschool[,c("name","pvM","ST57Q01","SC03Q01")]
study <- student.sub
names(study) <- c("Country","Math","HomeworkHours","Schoolplace")

p <- ggplot(data = study, aes(HomeworkHours,Math,colour=Schoolplace) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)
#ggsave(p,file="studytime.pdf",height=13,width=13)
p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") + facet_grid(SC03Q01~.) +xlim(0,30)
```

@@

```{r par_athomebar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
athome = student2012[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
athome = filter(athome, !is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
athome$parents = NA
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "Both"
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "Mother"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "Father"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
#athome$name <- reorder(athome$name, athome$PV1MATH, mean, decreasing=F)
#qplot(x = parents, data = athome) 
```

```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, facets = ~name, data = athome) + coord_flip() + ylab("Math Scores") + xlab("Family Structure") 
```

One of the variables the survey addressed was the occupation of each of the parents. There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g. home duties, retired)”, and no response. The distribution of the number of parents who selected each of the categories can be seen in the bar chart. The chart shows that father’s primarily have full-time jobs. However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms. With women making up a larger portion of the work-force, it is interesting to look at how the parent’s occupation relates to the student’s performance in mathematics, reading, and science.

```{r parentsjob, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
#student2012 <- student2012
student2012$ST15Q01 <- addNA(student2012$ST15Q01)
student2012$ST19Q01 <- addNA(student2012$ST19Q01)

>>>>

---
output: pdf_document
---
Seven Things That Affect Education Around the World
========================================================

@@

The OECD describes the PISA survey as ``the world's global metric for quality, equity and efficiency in school education" (citep[p. 3]{PISA2014a}). The goal of the PISA survey is to assess the workforce readiness of 15-year old students. Nearly 500,000 students were tested across the 65 countries and economies. Students are examined on how well they can apply the knowledge they learned in school to applications outside of school. Additional information about the students, parents, and schools is also collected. The students completed a questionnaire providing information about themselves, their homes, their schools, and a variety of psychological views regarding factors they believe affects their performance in school. Schools principals responded to a questionnaire covering the school system and learning experiences for students. In some countries, parents completed a questionnaire requesting information about their perceptions regarding the school system, expectations for their child, and their involvement in their child's schooling.

This paper will evaluate the role of several factors in which we see inequalities in mathematics, reading, and science outcomes. In section XXXgender, we will evaluate in what subjects we observe gender inequalities. In section XXX we will evaluate inequalities based on the household structure and occupation of the parents. In section XXXpossessions we will look at what home possessions are good predictors of inequalities. Lastly, in sections XXXschool, we will look at the role of school influences.

@@

```{r load_packages, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
library(dplyr)
library(stringr)
library(stringr)
library(reshape2)
library(ggplot2)
library(lubridate)
library(ggvis)
library(maps)
library(ggmap)
library(rworldmap)
library (grid)    
library (scales) 
```

```{r load_data, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
# This is the scaffolding to import the data
sets <- c("item", "parent", "school", "scoredItem", "student")

@@

world <- getMap(resolution = "low")
library(plyr)
world.polys <- extractPolygons(world)
detach("package:plyr")
```
### Teachers, computers, and libraries at school matter

### Gender gap in math is not universal but the reading gap is

### School start age affects performance

***We should decide on country names, to coordinate map info and the test data. This si some really ugly code to convert names to be the same between the two. Could be made cleaner, much cleaner!

Broken out by country we have computed the average math test score by age that the children started school. In additon, for each country we computed the median age that children in the study started school. There are differences in typical school start age between countries. Ireland has the youngest median start age - children head to school at age 4. Britain, and its past colonies, Australia, Canada and New Zealand, have a median start age of 5. For many other countries the median start age is 6, and for northern Europe, much of Asia and Brazil median start age is as late as 7. Some countries appear to have rigid starting ages, eg in Japan all children start at age 6.

```{r age, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]

@@

student2012.sub.age <- summarise(group_by(student2012.sub, name), 
                  math4 = mean(PV1MATH[ST06Q01 == 4], na.rm = T), 
                  math5 = mean(PV1MATH[ST06Q01 == 5], na.rm = T), 
                  math6 = mean(PV1MATH[ST06Q01 == 6], na.rm = T), 
                  math7 = mean(PV1MATH[ST06Q01 == 7], na.rm = T), 
                  math8 = mean(PV1MATH[ST06Q01 >  7], na.rm = T))
student2012.sub.age.m <- melt(student2012.sub.age)
student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
student2012.sub.agemedian <- summarise(group_by(student2012.sub, name), medianstart = median(ST06Q01, na.rm=T))
orderedbymedianage <- student2012.sub.agemedian$name[order(student2012.sub.agemedian$medianstart)]
student2012.sub.age.m.median <- left_join(student2012.sub.age.m, student2012.sub.agemedian)
student2012.sub.age.m.median$medianstart <- factor(student2012.sub.age.m.median$medianstart)
student2012.sub.age.m.median$name <- factor(student2012.sub.age.m.median$name, levels=orderedbymedianage)
qplot(variable, value, data=student2012.sub.age.m.median, xlab="Age at start", ylab="Math Score", 
      geom=c("point","smooth"), se=F, group=name, colour=medianstart) + 
 facet_wrap(~name, ncol=8) + theme_bw() + theme(legend.position="none")
# ggsave("school-start.pdf", width=14, height=14)
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  read4 = mean(PV1READ[ST06Q01 == 4], na.rm = T), 
#                  read5 = mean(PV1READ[ST06Q01 == 5], na.rm = T), 
#                  read6 = mean(PV1READ[ST06Q01 == 6], na.rm = T), 
#                  read7 = mean(PV1READ[ST06Q01 == 7], na.rm = T), 
#                  read8 = mean(PV1READ[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Reading Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  scie4 = mean(PV1SCIE[ST06Q01 == 4], na.rm = T), 
#                  scie5 = mean(PV1SCIE[ST06Q01 == 5], na.rm = T), 
#                  scie6 = mean(PV1SCIE[ST06Q01 == 6], na.rm = T), 
#                  scie7 = mean(PV1SCIE[ST06Q01 == 7], na.rm = T), 
#                  scie8 = mean(PV1SCIE[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Science Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 

@@

```{r maps, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
# Extract map polygons for modern world
world <- getMap(resolution = "low")
library(plyr)
world.polys <- extractPolygons(world)
detach("package:plyr")

student2012.sub.agemedian$medianstart <- factor(student2012.sub.agemedian$medianstart)
student2012.sub.map <- left_join(student2012.sub.agemedian, world.polys)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=medianstart)) +
  scale_fill_discrete("Median Start Age") +
  new_theme_empty + theme(legend.position="bottom")
#ggsave("agestartmap.pdf", width=12, height=7)
```
### Time spent out of school studying is important, but only up to a point
For each country we have provided loess smoothed curves of the relationship between mathematics and the time spent studying outside of school. The most prominent feature that we observe is the decreasing benefit of studying. For every country there is evidence of a substantial positive relationship for the students who spend less time studying. However, those students who continue to study for long hours see diminishing returns for their efforts. It is possible that there may be a selection issue occuring where more talented students spend less time studying as they complete their homework faster. However, the relationship shows those spending

```{r studying, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
student.luke <- student2012[,c("CNT","PV1MATH","PV1READ","PV1SCIE","ST57Q01")]
study.luke <- melt(student.luke,id=c("CNT","ST57Q01"))
studyplot <- ggplot(data = study.luke, aes(ST57Q01,value,colour=variable) )+facet_wrap(~CNT)+stat_smooth(se=F)
studyplot
```

```{r studyingbar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
studybar <- ggplot(data = study.luke, aes(ST57Q01) ) + geom_histogram(binwidth=4) + xlab("Hours spent studying")
studybar
```

@@

### subsection: Number of parents in the home
Looking at the number of parents in a household shows that student in two-parent household are on average performing better than student's not in a two-parent household. Students with just a mother at home do tend to perform better than students with only a father at home. Students with no parents at home are performing the worst, on average.

```{r par_athomebar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
athome = studentLuke[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
athome$parents = NA
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "Both"
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "Mother"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "Father"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
athome$name <- reorder(athome$name,athome$PV1MATH,mean)
pbarhome <- qplot(x = parents, data = athome) 
pbarhome

```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, facets = ~name, data = athome) + coord_flip() + ylab("Math") + xlab("Family Structure") 
phome
```

### subsection: Parent's Occupation Status
One of the variables the survey addressed was the occupation status of each of the parents. There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g. home duties, retired)”, and no response. The distribution of the number of parents who selected each of the categories can be seen in the bar chart. The chart shows that father’s primarily have full-time jobs. However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms. With women making up a larger portion of the work-force, it is interesting to look at how the parent’s occupation relates to the student’s performance in mathematics, reading, and science.

```{r parentsjob, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
studentLuke <- student2012
studentLuke$ST15Q01 <- addNA(studentLuke$ST15Q01)
studentLuke$ST19Q01 <- addNA(studentLuke$ST19Q01)
